Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages

### 2013-07-31 paper/main.tex
Draft IFL paper complete

\usepackage{amsmath}
\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usetikzlibrary{arrows,automata}
\WarningsOff

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

\titlebanner{Draft for submission to IFL 2013}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

@@

\begin{abstract}
Dependently-typed languages allow for much expressive types to be used within programs in turn allowing easier reasoning about the operation of such programs. Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.

@@

\begin{Verbatim}
import Cgi
import Effects

@@

main : IO ()
main = do runAction [initCGIState] sayHello
          pure ()
\end{Verbatim}
% =================================================

@@

We define a function of type:
\begin{Verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{Verbatim}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
\begin{Verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err $
\end{Verbatim}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

@@

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
\begin{Verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] $
\end{Verbatim}
In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
\begin{Verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{Verbatim}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
\begin{Verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{Verbatim}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
\begin{Verbatim}
executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
                 (List DBVal)) -> 
  Eff m [SQLITE ()] (Either String ResultSet)
\end{Verbatim}
% =================================================
\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

@@

\begin{Verbatim}
'SELECT `name`, `address` FROM `staff` LIMIT ' + page + '
                                                     , 5'
\end{Verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

@@

\subsection{Form Construction}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
\begin{Verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{Verbatim}
Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
\begin{Verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{Verbatim}
%
In order to specify a form, we once again use \texttt{Effects}. By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.

@@

To store a reference to a handler function, we use the \texttt{HandlerFn} type:
\begin{Verbatim}
HandlerFn : Type
HandlerFn = (ft ** (mkHandlerFn ft, String))
\end{Verbatim}
%
The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function. We also store a unique string identifier, which is used to serialise a reference to the handler function. We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.

\begin{Verbatim}
HandlerList : Type
HandlerList = List HandlerFn
\end{Verbatim}
To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
\begin{Verbatim}
  data FnElem : mkHandlerFn ((reverse G), E) -> 
                HandlerList -> Type where

@@

Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page. Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}. 
\subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations. 
\begin{Verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{Verbatim}
We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled. If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
\begin{Verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{Verbatim}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
\begin{Verbatim}
newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers
\end{Verbatim}
The form consists of two text boxes: one for the title of the thread, and one for the content of the first post. Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value. The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
\begin{Verbatim}
handlers : HandlerList
handlers = 
  [(([FormString, FormString], 
     [CgiEffect, SessionEffect, SqliteEffect]) ** 
     (handleRegisterForm, "handleRegisterForm")),

   (([FormString, FormString], 
     [CgiEffect, SessionEffect, SqliteEffect]) ** 
     (handleNewThread, "handleNewThread")),

@@

           (output \$ "<tr><td>

@@

        lift' (output \$ "Welcome, " ++ name)

@@

      lift' (output \$ "Error: " ++ err)

@@

\section{Related Work}

@@

\section{Conclusions}
Dependently-typed languages have, for a long time, remained relatively unused outside of academia. Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.

@@

Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations. Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly. By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly. Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.

@@

>>>>

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
\usetikzlibrary{arrows,automata}
%\WarningsOff

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\newcommand{\idata}{\textsf{iData}}
\newcommand{\itasks}{\textsf{iTasks}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

\titlebanner{}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

@@

\begin{abstract}
Dependently-typed languages allow much expressive types to be used within programs, in turn allowing easier reasoning about the operation of such programs. Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.

@@

\begin{Verbatim}
module Main
import Cgi

@@

main : IO ()
main = runCGI [initCGIState] sayHello
\end{Verbatim}
% =================================================

@@

We define a function of type:
\begin{Verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{Verbatim}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
\begin{Verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err 
\end{Verbatim}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

@@

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
\begin{Verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] 
\end{Verbatim}
In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
\begin{Verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{Verbatim}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
\begin{Verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{Verbatim}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
\begin{Verbatim}
executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
                 (List DBVal)) -> 
  Eff m [SQLITE ()] (Either String ResultSet)
\end{Verbatim}
% =================================================
\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

@@

\begin{Verbatim}
'SELECT `name`, `address` FROM `staff` LIMIT ' + 
       page + ', 5';
\end{Verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

@@

We begin with a simple example of a form which requests a user's name, and echoes it back. Firstly, we define a form handler which echoes back a string provided by the form handler. It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.

\begin{Verbatim}
sayHello : Maybe String -> 
           FormHandler [CGI (InitialisedCGI TaskRunning)]
sayHello (Just name) = output ("Hello, " ++ name ++ "!")
sayHello _ = output "Error!"
\end{Verbatim}

We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:

\begin{Verbatim}
handlers : HandlerList
handlers = [(handler args=[FormString], 
                     effects=[CgiEffect], 
                     fn=sayHello, 
                     name="sayHello")
           ]
\end{Verbatim}

We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.

\begin{Verbatim}
showHelloForm : UserForm
showHelloForm = do
  addTextBox "Name" FormString Nothing
  useEffects [CgiEffect]
  addSubmit sayHello handlers
\end{Verbatim}

Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.

\begin{Verbatim}
cgiHello : CGIProg [] ()
cgiHello = do
  handler_set <- isHandlerSet
  if handler_set then do
    handleForm handlers
    return ()
  else do
    addForm "nameform" "helloform" showHelloForm
    return ()

main : IO ()
main = runCGI [initCGIState] cgiHello
\end{Verbatim}
When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user. Upon submission of the form, the form handler will be invoked, and the name will be used in the output.

In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation of the form-handling system: namely, the EDSL which allows the creation of forms, and the handling code which deserialises the data and passes it to the user-specified handler function.
\subsection{Form Construction}
\label{formcons}
Each form element is specified to hold a particular type of data, which is then passed directly as that particular type, assuming that the correct type of data is specified by the user, to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.
\begin{Verbatim}
data FormTy = FormString
            | FormInt
            | FormBool
            | FormFloat
            | FormList FormTy 
\end{Verbatim}
Recalling that types in \idris{} are first-class, we may use this to convert between abstract and concrete representations of allowed form types:
\begin{Verbatim}
interpFormTy : FormTy -> Type
interpFormTy FormString = String
interpFormTy FormInt = Int
interpFormTy FormBool = Bool
interpFormTy FormFloat = Float
interpFormTy (FormList a) = List (interpFormTy a)
\end{Verbatim}
%
In order to specify a form, we once again use \texttt{Effects}. By recording the type of each form element as it is added in the type of the form, we may statically ensure that the user-supplied handler function is of the correct type to handle the data supplied by the form: the specification of an incompatible handler will result in a compile-time type error.

@@

To store a reference to a handler function, we use the \texttt{HandlerFn} type:
\begin{Verbatim}
HandlerFn : Type
HandlerFn = (ft ** (mkHandlerFn ft, String))
\end{Verbatim}
%
The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function. We also store a unique string identifier, which is used to serialise a reference to the handler function.

In order to abstract away from this implementation detail, we make use of \idris{} syntax rewriting rules. This allows us to define the following syntax rewrite rule:
\begin{Verbatim}
syntax 
  "handler args=" [args] ", effects=" [effs] ", fn=" [fn] 
  ", name=" [name] = ((args, effs) ** (fn, name))
\end{Verbatim}
We may then define handlers in a more intuitive fashion, without being concerned with the implementation details. This allows us to write a handler with one String argument, making use of the CGI effect, associated with the \texttt{sayHello} handler function as follows:
\begin{Verbatim}
handler args=[FormString], 
        effects=[CgiEffect], 
        fn=sayHello, 
        name="sayHello"
\end{Verbatim}

We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.
\begin{Verbatim}
HandlerList : Type
HandlerList = List HandlerFn
\end{Verbatim}
To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
\begin{Verbatim}
  data FnElem : mkHandlerFn ((reverse G), E) -> 
                HandlerList -> Type where

@@

Secondly, we use a GET variable, \texttt{action}, to indicate which page of the message board should be displayed, and pattern-match on these to call the appropriate function which displays the page. Some pages, such as the page which shows all of the posts in a thread, require a second argument, \texttt{thread\_id}. 
\subsection{Handling requests}
The entry point to any CGI application is the \texttt{main} function. From here, we run the remainder of the program through a call to \texttt{runCGI}, which we initialise with empty initial environments for the CGI, Session and SQLite effects, so they may be used in further computations. 
\begin{Verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{Verbatim}
We specify a function, \texttt{handleRequest}, which firstly calls \texttt{isHandlerSet} to determine whether submitted form data must be handled. If so, then the form handling routine is called, which executes the corresponding handler function as specified in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
\begin{Verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{Verbatim}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.
\begin{Verbatim}
newThreadForm : UserForm
newThreadForm = do
  addTextBox "Title" FormString Nothing
  addTextBox "Post Content" FormString Nothing 
  useEffects [CgiEffect, SessionEffect, SqliteEffect]
  addSubmit handleNewThread handlers
\end{Verbatim}
The form consists of two text boxes: one for the title of the thread, and one for the content of the first post. Both are of type \texttt{String}, as denoted by the \texttt{FormString} argument, and both have no default value. The handler function may make use of the \texttt{CGI}, \texttt{SESSION} and \texttt{SQLITE} effects, and the handler function is specified as \texttt{handleNewThread}. The \texttt{handlers} argument refers to the list of form handlers, and is of the following form:
\begin{Verbatim}
handlers : HandlerList
handlers = [
 (handler args=[FormString, FormString], 
          effects=[CgiEffect, SessionEffect, SqliteEffect], 
          fn=handleRegisterForm, 
          name="handleRegisterForm"),

 (handler args=[FormString, FormString], 
          effects=[CgiEffect, SessionEffect, SqliteEffect], 
          fn=handleNewThread, 
          name="handleNewThread"),

@@

           (output $ "<tr><td>

@@

        lift' (output $ "Welcome, " ++ name)

@@

      lift' (output $ "Error: " ++ err)

@@

\section{Related Work}
\citet{meijer:cgi} implemented a CGI library which was among the first libraries to handle web scripting monadically, and allows the user to implement application logic without having to consider the low-level details such as parsing in CGI data from the environment, or printing headers to the remote browser. The library also provides support for cookies and basic form handling.

\citet{thiemann:wash} further adds the notion of a CGI Session for maintaining state, and provides more sophisticated form-handling methods. In particular, callbacks may be associated with submit buttons, with nameless representations for form inputs. Due to the unavailability of full dependent types in Haskell, however, this implementation does not statically verify the suitability of the callback function for the form inputs. Both implementations of the CGI library, being built upon monads, mean that the use of additional effects such as database access is achieved either through monad transformers or the ability to perform arbitrary IO operations. Both of these approaches are limited---the former does not scale well to multiple effects, and the latter allows for the introduction of errors by allowing the violation of resource usage protocols.

\citet{plasmeijer:idata} describe an alternative approach to type-safe form handling which does not require the use of dependent types through the \textit{i}nteractive \textit{Data}, or \idata{} abstraction. Instead of processing being triggered by form submission, as in the approach described in this paper, applications created in the \idata{} toolkit are instead edit-driven. This means that upon a component being edited, a computation occurs, given the state of the current form. This is then saved to be used in future computations. Should a user enter invalid data, for example by entering text in a field designated for integers, the change will be reverted. This is demonstrated practically through the use of \idata{} to implement a conference management system \cite{plasmeijer:cms}.

The concept of \idata{} is taken further by the introduction of \itasks{} \cite{plasmeijer:itasks}, which make use of a workflow system to allow multiple \idata{} forms to interact with one another. This is achieved through the use of high-level combinators which allow the implementation of concepts such as recurseion, sequence and choice in a scalable fashion. These high-level abstractions are elegant, but the style and syntax differ substantially from a traditional web application. Our approach takes the concept of type-safe input handling and uses it in a more traditional fashion, whilst retaining the type-retention guarantees afforded by \idata{} elements.

Ur\/Web \cite{urweb} is a library for the Ur language which makes use of dependent types to enable the generation of provably-correct and secure queries and DOM code, without requiring overly complex types or proofs. Through the use of the Ur\/Web DSL, developers may generate well-typed code from records. By statically verifying the \textit{code generator}, it is therefore possible to assert that the generated code is secure. Such ideas are extremely interesting for the implementation of object-relational mapping, which we hope to implement in the near future.

WebDSL \cite{webdsl} is a domain-specific language written primarily to introduce new abstractions which aim to reduce the amount of boilerplate code that must be written and maintained by developeres. WebDSL is built on top of Java, which often includes a large amount of redundant code such as accessor and mutator functions within entity classes. The Java Persistence API (JPA) \cite{jpa} provides an object-relational mapping through the use of Java 5 annotations, which may then be used to construct database tables. These annotations soon become complex, however, and coupled with redundant boilerplate code, data model declarations may soon become unwieldy. Through the use of WebDSL, these data model declarations can be much more compactly declared, and elaborated into Java code by parsing the data-modelling DSL into an abstract syntax tree, applying rewrite rules, and pretty-printing. WebDSL also applies similar concepts to implement a \textit{template system} for such objects, which allows the data to be used in code generation. We look to implement many of these ideas, but as DSLs embedded within the IdrisWeb framework, as with the form construction EDSL.

@@

\section{Conclusions}
Dependently-typed languages have, remained relatively unused for practical purposes outside of academia. Advances in embedded domain-specific languages have allowed many of the advantages of such languages, including the additional guarantees given by the successful compilation of a program (assuming the use of specific enough types) to be adopted without the additional proofs traditionally required by users of such languages.

@@

Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations. Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly. By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly. Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.

@@

Since at this stage we have concentrated on the use of dependent types for enforcing resource usage protocols and type-safe form handling, we currently handle the generation of HTML in an unstructured manner. Future work will entail a DOM library to facilitate the generation and manipulation of HTML, in turn giving stronger guarantees about its correctness. Other planned features include a template system, allowing for web pages to be automatically generated from data, and an object-relational mapping system which will allow users to manipulate records which can then be automatically written to the database, instead of having to update tables manually via SQL queries.

Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used in compilation. Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures. If data structures within the program do not conform to the schema, then the program will not type-check and as such will produce a compile-time error. This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step. Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness. Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.
